
\documentclass{article}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%TCIDATA{OutputFilter=LATEX.DLL}
%TCIDATA{Version=5.00.0.2552}
%TCIDATA{<META NAME="SaveForMode" CONTENT="1">}
%TCIDATA{Created=Monday, November 27, 2006 12:19:02}
%TCIDATA{LastRevised=Monday, December 03, 2007 13:42:20}
%TCIDATA{<META NAME="GraphicsSave" CONTENT="32">}
%TCIDATA{<META NAME="DocumentShell" CONTENT="Scientific Notebook\Blank Document">}
%TCIDATA{CSTFile=Math with theorems suppressed.cst}
%TCIDATA{PageSetup=72,72,72,72,0}
%TCIDATA{AllPages=
%F=36,\PARA{038<p type="texpara" tag="Body Text" >\hfill \thepage}
%}


\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{algorithm}[theorem]{Algorithm}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\newenvironment{proof}[1][Proof]{\noindent\textbf{#1.} }{\ \rule{0.5em}{0.5em}}
\input{tcilatex}

\begin{document}


\textbf{Lecture notes 09}

\textbf{EC 322}

\textbf{Dr. David Reinstein }

\textbf{27 Nov. 2006}\bigskip

\textbf{Warning}:\ These notes are subject to error. \ Not for opthamalmic
use.

\textbf{VII. 28 Nov. -- 1 Dec.}

Dynamic Games with Imperfect/Incomplete information;

Signaling games; Catch-Up, Applications of Previous Topics, Review for
Midterm

\textbf{Reading}

Osborne: (\textit{10.1-10.4 if you haven't yet}), 10.5, 10.7; Review
previous readings and articles for midterm

\bigskip

\section{Extensive game with imperfect information and chance moves}

We have already seen this (remember the games with some dynamics and some
simultaneity)

\begin{itemize}
\item Now we will allow simultaneity at any point in the game

\item Now we will include a new player:\ `chance' (AKA nature)
\end{itemize}

\bigskip

However:

\begin{quote}
As we consider progressively richer games, we progressively strengthen the
equilibrium concept, in order to rule out implausible equilibria in the
richer games that would survive if we applied equilibrium concepts suitable
for simpler games (Gibbons, p. 173)
\end{quote}

Nash Equilibrium (NE)$\ \mathbf{\rightarrow }$ subgame-perfect NE (SPNE) $%
\mathbf{\rightarrow }$ Bayesian NE $\mathbf{\rightarrow }$ Weak Sequential
Equilibrium

Static games of perfect information $\mathbf{\rightarrow }$ dynamic games of
perfect (or `some imperfect') information $\mathbf{\rightarrow }$ static
games of imperfect information $\mathbf{\rightarrow }$

dynamic games of imperfect information

\textbf{Figure 316.1: `Poker'}

\[
\FRAME{itbphF}{5in}{4in}{0in}{}{}{Figure}{\special{language "Scientific
Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display "USEDEF";valid_file
"T";width 5in;height 4in;depth 0in;original-width 4.6799in;original-height
3.584in;cropleft "0";croptop "1";cropright "1";cropbottom "0";tempfilename
'JSH1WS00.wmf';tempfile-properties "XPR";}} 
\]

\bigskip

\textbf{Exercise 316.1}

\[
\FRAME{itbphF}{6.5164in}{5.2129in}{0pt}{}{}{Figure}{\special{language
"Scientific Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display
"USEDEF";valid_file "T";width 6.5164in;height 5.2129in;depth
0pt;original-width 7.32in;original-height 5.0087in;cropleft "0";croptop
"1";cropright "1";cropbottom "0";tempfilename
'JSH1WS01.bmp';tempfile-properties "XP";}} 
\]

\bigskip

\textbf{Figure 317.1: Challenger who may be ready or unready and Incumbent}

\begin{equation}
\FRAME{itbphFU}{5.0004in}{4in}{0in}{\Qcb{Fig. 317.1}}{}{Figure}{\special%
{language "Scientific Word";type "GRAPHIC";maintain-aspect-ratio
TRUE;display "USEDEF";valid_file "T";width 5.0004in;height 4in;depth
0in;original-width 5.6886in;original-height 3.5201in;cropleft "0";croptop
"1";cropright "1";cropbottom "0";tempfilename
'JSH1WS02.wmf';tempfile-properties "XPR";}}
\end{equation}

\bigskip

`\textbf{Strategy':}

We already defined this. \ Assigns an action (or a mix -- we will get to
this later)\ for each information set.

\bigskip

\textbf{Nash Equilibrium }

As before.

\bigskip

\subsection{Figure 316.1: `Poker'}

\begin{equation}
\FRAME{itbphF}{5in}{4in}{0in}{}{}{Figure}{\special{language "Scientific
Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display "USEDEF";valid_file
"T";width 5in;height 4in;depth 0in;original-width 4.6799in;original-height
3.584in;cropleft "0";croptop "1";cropright "1";cropbottom "0";tempfilename
'JSH1WT03.wmf';tempfile-properties "XPR";}}
\end{equation}

\bigskip

Payoffs expressed as EU payoffs (over both types)

\begin{tabular}{ccc}
& \textbf{Pass} & \textbf{Meet} \\ 
\textbf{R,R} & \underline{1},-1 & 0,\underline{0} \\ 
\textbf{R,S} & 0,\underline{0} & \underline{$\frac{1}{2}$}$,-\frac{1}{2}$ \\ 
\textbf{S,R} & \underline{1},-1 & -$\frac{1}{2},$\underline{$\frac{1}{2}$}
\\ 
\textbf{S,S} & 0,\underline{0} & 0,\underline{0}%
\end{tabular}

\bigskip

S,S dominated by $P(R,R)\otimes \frac{1}{2}+P(R,S)\otimes \frac{1}{2}$

$\mathbf{\Longrightarrow }S,S$ not used in any NE

\bigskip

\bigskip 
\begin{tabular}{ccc}
& \textbf{Pass} & \textbf{Meet} \\ 
\textbf{R,R} & \underline{1},-1 & 0,\underline{0} \\ 
\textbf{R,S} & 0,\underline{0} & \underline{$\frac{1}{2}$}$,-\frac{1}{2}$ \\ 
\textbf{S,R} & \underline{1},-1 & $-\frac{1}{2},$\underline{$\frac{1}{2}$}%
\end{tabular}

No equilibrium in pure strategies.

$\bigskip $

$\mathbf{\Longrightarrow }\Pr (meet)>0$

$\mathbf{\Longrightarrow }S,R$ not a BR

\bigskip

--If Meet than R,R better

\bigskip

--If Pass than R,R as good

$\bigskip $

$\mathbf{\Longrightarrow }$So don't put any probability on $S,R$; would do
better by moving this to $R,R$

$\bigskip $

$\mathbf{\Longrightarrow }$Mix between R,R and R,S

\bigskip

Which I claim was the intuitive result. \ 

\bigskip

You will always raise if you have the high cards (there is no way to do
better).

You will sometimes raise if you have the low cards (i.e., bluff).

\bigskip

BNE:\ $(P(R,R)\otimes \frac{1}{3}+P(R,S)\otimes \frac{2}{3},P(p)\otimes 
\frac{1}{3}+P_{m}\otimes \frac{2}{3})$

Bluffs with probability $\frac{1}{3}$

`always calls bluff' with probability $\frac{2}{3}$

\bigskip

\subsection{Exercise 320.2:\ Game as simultaneous, dynamic, dynamic with
imperfect observation}

\[
\begin{tabular}{ccc}
& \textbf{x} & \textbf{y} \\ 
\textbf{x} & 3,\underline{2} & 1,1 \\ 
\textbf{y} & \underline{4},3 & \underline{2,4}%
\end{tabular}%
\]

This is a game with `1st mover advantage'. \ Let's say it is not
simultaneous but 1 moves first

\FRAME{ftbphF}{5in}{4in}{0pt}{}{}{Figure}{\special{language "Scientific
Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display "USEDEF";valid_file
"T";width 5in;height 4in;depth 0pt;original-width 4.3196in;original-height
3.4321in;cropleft "0";croptop "1";cropright "1";cropbottom "0";tempfilename
'JSH1WT04.wmf';tempfile-properties "XPR";}}

In the NE of this dynamic game player 1 now does better than under the
simultaneous game.

\newline
\textit{Q: Recall a game with first-mover disadvantage?}

\bigskip

A. Rocks, Paper, Scissors, for example.

A. Also, Stackleberg game.

\bigskip

\textbf{Now, suppose there is an `imperfect signal'. }

Player 2 sees what player 1 does but with a probability $\varepsilon $
mispercieves this

\bigskip

$\Pr (signal~is~correct)=1-\varepsilon $

Assume $0\leq \varepsilon \leq \frac{1}{4}$

\bigskip

Example:\ you can work or be lazy. \ There is some probability your boss
will fail to detect this (in either direction). \ 

\bigskip

Drawing this as a game tree (extensive form):

\bigskip (Fig. 321.1)%
\begin{equation}
\FRAME{itbphF}{5in}{4in}{0pt}{}{}{Figure}{\special{language "Scientific
Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display "USEDEF";valid_file
"T";width 5in;height 4in;depth 0pt;original-width 5.5201in;original-height
3.7758in;cropleft "0";croptop "1";cropright "1";cropbottom "0";tempfilename
'JSH1WT05.wmf';tempfile-properties "XPR";}}
\end{equation}

\textbf{In the strategic/normal form (matrix), written in terms of EU:}

\[
\begin{tabular}{ccccc}
& \textbf{XX} & \textbf{XY} & \textbf{YX} & \textbf{YY } \\ 
& X always & trusting & paranoid & Y always \\ 
\textbf{X} & $3,\underline{2}$ & $2-2\varepsilon ,2-\varepsilon $ & $%
1+2\varepsilon ,1+\varepsilon $ & $1,1$ \\ 
\textbf{Y} & $\underline{4},3$ & \underline{$2+2\varepsilon $}$%
,4-\varepsilon $ & \underline{$4-2\varepsilon $}$,3+\varepsilon $ & $%
\underline{2},\underline{4}$%
\end{tabular}%
\]

Noting, e.g., $(1-\varepsilon )3+1\varepsilon =2-2\varepsilon $

\bigskip

$\mathbf{\Longrightarrow \{}Y,YY\}$ is unique PS-NE strategy profile

\bigskip

Which is the same as in the simultaneous game. \ With only a little
uncertainty about the previous move!

\bigskip

\bigskip

\subsection{Example 322.1: \ entry game, Challenger who may be ready or
unready and Incumbent}

\textbf{Extensive form :\bigskip }

\begin{equation}
\FRAME{itbphFU}{5.0004in}{4in}{0in}{\Qcb{Fig. 317.1}}{}{Figure}{\special%
{language "Scientific Word";type "GRAPHIC";maintain-aspect-ratio
TRUE;display "USEDEF";valid_file "T";width 5.0004in;height 4in;depth
0in;original-width 5.6886in;original-height 3.5201in;cropleft "0";croptop
"1";cropright "1";cropbottom "0";tempfilename
'JSH1WT06.wmf';tempfile-properties "XPR";}}
\end{equation}

Note:\ no proper subgames

(remember, a subgame cannot cut an information set)

So SPNE is the same as NE here.

\bigskip

\textbf{Strategic form:}

\begin{equation}
\begin{tabular}{ccc}
& \textbf{A} & \textbf{F} \\ 
\textbf{R} & 3,3 & 1,1 \\ 
\textbf{U} & \underline{\textit{4}},\underline{3} & 0,3 \\ 
\textbf{O} & 2,4 & \underline{\textit{2,4}}%
\end{tabular}%
\end{equation}

\bigskip

There are two NE here: \{Unready,Acquiesce\} and \{Out,Fight\}?.\ 

Are they both reasonable?

\bigskip

Subgame perfection doesn't eliminate either (no proper subgames).

\bigskip

But we know it is suboptimal for the incumbent to fight once the challenger
has entered, whether or not the challenger is prepared. \ So `fight'
wouldn't seem to be a good strategy ever, and thus \{Out,Fight\} depends on
an incredible threat.

\bigskip

\textit{We could add a requirement:}

\textbf{Optimal at each information set:} Each player's strategy is optimal
(for at least some node in the information set) whenever she has to move.

I.e., each player's strategy must be optimal at each of her information sets.

\bigskip

The strategy `Fight' is not optimal at each information set here -- it is
not optimal for the incumbent to fight at the incumbent's information set
(when he has the move, i.e., after some form of entry), niether whether he
is at the node following Unready nor whether he is at the node following
Ready

\bigskip

Thus, only the NE strategy profile \{Unready,Acquiesce\} meets the
requirement of optimality at each information set.

\subsection{Variant of this game:}

\textbf{Extensive form:}

(Figure 323.1)\bigskip 
\[
\FRAME{itbphFU}{5.0004in}{4in}{0in}{\Qcb{Fig 323.1}}{}{Figure}{\special%
{language "Scientific Word";type "GRAPHIC";maintain-aspect-ratio
TRUE;display "USEDEF";valid_file "T";width 5.0004in;height 4in;depth
0in;original-width 5.4803in;original-height 3.4396in;cropleft "0";croptop
"1";cropright "1";cropbottom "0";tempfilename
'JSH1WT07.wmf';tempfile-properties "XPR";}} 
\]

Strategic form:

\[
\begin{tabular}{ccc}
& \textbf{A} & \textbf{F} \\ 
\textbf{R} & 3,2 & 1,1 \\ 
\textbf{U} & \underline{\textit{4}}\underline{\textit{,2}} & 0,3 \\ 
\textbf{O} & 2,4 & \underline{\textit{2,4}}%
\end{tabular}%
\]

There are again 2 pure strategy NE here. \ 

But are they both reasonable?

\bigskip

Now the incumbent prefers to fight if the challenger enters unready, but
prefers to acquiese if the challenger enters ready. \ So there is no clear
optimal move at this information set -- it depends whether the challenger
prepared.

\bigskip

So, is Out, Fight a reasonable equilibrium?

Perhaps if the incumbent thinks `any challenger who enters will not prepare.'

At this node (following Unready) the action `Fight' is optimal.

Thus the requirement of `optimality at each information set' does not rule
anything out here.

\bigskip

Note that whether fight is optimal depends on the incumbent's \textbf{%
beliefs }(about out-of-equilibrium play)!

\bigskip

\subsection{Definitions: Beliefs, sequential equilibrium, assesments}

\textbf{Rem: }

\bigskip

In the static game beliefs were `correct.'

\bigskip

For extensive game this is less clearly defined -- some nodes are never
reached, so there is no obvious `correct' belief!

\bigskip

$\mathbf{\Longrightarrow }$\textit{We will define equilibrium in terms of
pairs: Strategy profiles \& Collections of beliefs }

\bigskip

\textbf{Beliefs }

At an information set with more than one history (element/node) the player
who has the move `forms a belief over the histories' (previous moves).

At each information set a \textit{belief }assigns a probability to each
possible history,

...or assigns a probability distribution over all the possible histories (in
a non-discrete game)..

\bigskip

\textbf{Belief System}

I.e., \ a \textit{belief system }is a collection of \textit{beliefs}, one
for each information set of each player.

\bigskip

\textbf{Behavioral strategies}

Can assign probabilities over actions at a particular node (not just over
complete pure strategies)

\bigskip

\emph{Example:}

Say there is a dynamic game where you can ask someone out on a date or not,
then they can say yes or no, then, if they say `yes', you choose whether to
go to the movies or out to dinner, and if they say `no' you choose whether
to go out with friends for a laugh or go home and have a good cry.

\bigskip

If you decide, `\textit{I will flip a coin...}

\textit{Heads: I ask him/her out, if they say `yes' we do out to dinner, if
they say no I go home and have a good cry' (i.e., \{Ask,Dinner,Cry\})}

\textit{Tails: I ask him/her out, if they say `yes' we do out to the movies,
if they say no I go home and have a good cry' (i.e., \{Ask,Movies,Cry\})}'

... this is a \textit{mixed strategy.}

I.e., $\{P(\{Ask,Dinner,Cry\})\otimes \frac{1}{2}+P(\{Ask,Movies,Cry\})%
\otimes \frac{1}{2}\}$

\bigskip

On the other hand if you decide:

\textit{`I\ will ask him or her out, then if he/she says `yes' I\ will then
flip a coin, }

\textit{...Heads: we go to the movies, Tails:\ we go to Dinner ...}

\textit{and if he/she says `no' I\ will go home and have a good cry'}

This is a \textit{behavioral }strategy.

I.e., $\{Ask,P(Dinner)\otimes \frac{1}{2}+P(Movies)\otimes \frac{1}{2},Cry\}$

\textit{Example II:}

\bigskip 
\begin{equation}
\FRAME{itbphF}{5in}{4in}{0in}{}{}{Figure}{\special{language "Scientific
Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display "USEDEF";valid_file
"T";width 5in;height 4in;depth 0in;original-width 4.264in;original-height
3.1922in;cropleft "0";croptop "1";cropright "1";cropbottom "0";tempfilename
'JSH1WT08.wmf';tempfile-properties "XPR";}}
\end{equation}

E.g., in the locking door game above, a mixed strategy for the second mover
(thief) is:

Play \{check,Check$^{\prime }$)\ with probability $\frac{1}{2}$ and
\{check,Don't$^{\prime }$\} with probability $\frac{1}{2}$.

\bigskip

While a \textit{behavioral strategy }is to play $\{check,P_{Check^{\prime
}}\otimes \frac{1}{2}+P_{Don^{\prime }t^{\prime }}\otimes \frac{1}{2}\}$

I.e., to play `check' for sure if player 1 plays Lock and to mix between
Check$^{\prime }$ and Don't$^{\prime }$ (with probabilities of $\frac{1}{2}$
each) if player 1 plays Don't.

These are in effect the same thing in all the cases we will encounter. But
it will be a notational convenience.

\textbf{Note:\ }For lack of time, these lecture notes do not look at any
examples of mixing/behavioral strategies.

\bigskip

\textbf{Def: }In general, a \textbf{behavioral strategy} of player $i$ in an
extensive game is a function that assigns , to each of $i$'s information
sets I$_{i}$, a probability distribution over the actions in A(I$_{i}$) ...
\ 

...Of course these probabilities could be zero or one (pure strategies).

\bigskip

\textbf{Assessment}

An \textit{assessment} pairs a profile of behavioral strategies and a belief
system.

\bigskip

\subsection{Equilibrium Concept:\ Perfect Bayesian Equilibrium -- PBE  (also
see Weak Sequential Equilibrium )}

Perfect Bayesian Equilibrium  \textbf{\footnote{%
There are small difference between PBE\ and WSE but I will not discuss them
here.}}

An assesment is a \textit{PBE }if it meets two requirements

\bigskip

\textit{Requirement 1}\textbf{\ -- Sequentially rationality:} Each strategy
is optimal for each player when he has the move given his belief (and the
other players' subsequent prescribed `continuation' strategies)

\bigskip

\textit{Requirement 2 -- }\textbf{\ Weak Consistency of Beliefs with
Strategies: }$\ $Each player's belief is consistent with the strategy
profile (this will be made more specific later)..

\bigskip

\textit{Sequentially rationality} rules out off-the-equilibrium-path
behavior that is `suboptimal' (given beliefs and the other player's
subsequent strategies)

\pagebreak

\begin{equation}
\FRAME{itbphFU}{5.0004in}{4in}{0in}{\Qcb{Fig. 326.1}}{}{Figure}{\special%
{language "Scientific Word";type "GRAPHIC";maintain-aspect-ratio
TRUE;display "USEDEF";valid_file "T";width 5.0004in;height 4in;depth
0in;original-width 5.3599in;original-height 4.4317in;cropleft "0";croptop
"1";cropright "1";cropbottom "0";tempfilename
'JSH1WT09.wmf';tempfile-properties "XPR";}}
\end{equation}

\begin{itemize}
\item Above, suppose that player 1 plays the strategy highlighted in blue
(player 2's moves are in red)

\item Given player 1's strategy what should 2 suppose if he `gets the move'
(i.e., arrives at information set \{C,D\}?

\item Let's say (we are trying this out), that if 1 decides to deviate from
his strategy (E) 2 supposes that 1 plays C with probability $\frac{2}{3}$
and D with probability $\frac{2}{3}$.

\item I.e., 2's \textit{beliefs} at this information set assign the
probabilities $\frac{2}{3}$ and $\frac{1}{3}$ to C and D, respectively, as
shown in brackets.

\item We assume that 2 will believe that, even after making this `mistake'
or `deviation' 1 will subsequently play rationally, i.e., play J if she gets
to the node following C,F.
\end{itemize}

\bigskip

\textit{Given the assumptions above, }\textbf{what is 2's `best'
(sequentially rational)\ action at the information set \{C,D\}?}

Let's consider the EU\ payoffs to each move:

\begin{itemize}
\item If 1 played C, 2's best move is G, yielding payoff 1. \ If 1 played D,
G yields 0.

\item If 1 played D, 2's best move is F, yielding payoff 1. If 1 played C, F
yields 0.

\item So which action is better depends on the probability 1 played C versus
D

\item (2 believes that) C occured with probability $\frac{2}{3}$ and D with
probability $\frac{1}{3}$

\item Thus 2's EU payoffs to playing G are:
\end{itemize}

\begin{eqnarray*}
EU_{2}(G) &=&\Pr (C)\times U_{2}(\{C;G\})+\Pr (D)\times U_{2}(\{D;G\}) \\
&=&\frac{2}{3}\times 1+\frac{1}{3}\times 0=\frac{2}{3}
\end{eqnarray*}

\begin{itemize}
\item On the other hand, 2's EU payoffs to playing F are:
\end{itemize}

\begin{eqnarray*}
EU_{2}(F) &=&\Pr (C)\times U_{2}(\{C;F\})+\Pr (D)\times U_{2}(\{D;F\}) \\
&=&\frac{2}{3}\times 0+\frac{1}{3}\times 1=\frac{1}{3}
\end{eqnarray*}

So, given the assumptions above, given 2's beliefs and 1's optimal strategy
(for the rest of the game), if 2 is at the information set \{C,D\}, his best
move is G. \ G is the `sequentially rational' action given these beliefs and
1's strategy.

\bigskip

\textbf{Hold on, you! If 1 is choosing E\ anyway, why do we need to worry
about this?}

Ok, so we are only `considering' that 1 may play E. \ 

1 must ask \textit{`is playing E the best?'} \ `\textit{Will playing C or D
do better (in EU terms)?'}

To get the answer to this question, 2 needs to ask `\textit{if I\ play C (or
D), what will player 2 do?}'

But what player 2 does at \{C,D\} will depend on what player 2 \textit{%
believes }1 played -- C or D.

So, in order to decide the the EU payoffs to playing C (or D), to compare
this to playing E, 1 needs to know (or assign) what player 2's \textit{%
beliefs }are at \{C,D\}.

\pagebreak \bigskip

\subsubsection{Weak Consistency of Beliefs with Strategies (Requirement 2)}

`In a steady state, each player's belief must be correct' -- Osborne

\textbf{Def. }$\ $An information set is \textbf{on the equilibrium path }if
it will be reached \textit{with some positive probability} if the game is
played according to the `equilibrium strategies' (using whichever
equilibrium concept we chose).

Otherwise an information set is \textbf{off the equilibrium path}

\bigskip

\textbf{Consistency of beliefs at information sets on the equilibrium path}%
\textit{:}

`At information sets \textit{on the equilibrium path}, beliefs are
determined by Bayes rule (i.e., correct statistical inference)\ and the
players' equilibrium strategies.\footnote{%
PBE also requires: Consistency of beliefs at information sets off the
equilibrium path
\par
`At information sets off the equilibrium path, beliefs are determined by
Bayes' rule and the players' equilibrium strategies \textit{where possible'}
\par
{}}

\bigskip

\textbf{Bayes rule} for conditional probabilities in this case:

\[
\Pr (h^{\ast }|I_{i}\text{ reached, behavioral strategy }\beta )=\frac{\Pr
(h^{\ast }|\beta )}{\sum\limits_{h\in I_{i}}\Pr (h|\beta )} 
\]

A player finds herself at information set $I_{i}$. \ Knowing all players'
behavioral strategies $\beta $. \ 

What probability should she assign that $h^{\ast }$ has been played? \ 

She should take the ratio of

...the (unconditional)\ probability that $h^{\ast }$ is played under the
behavioral strategy $\beta $ \ 

...to the probability we \textit{get }to the information set $I_{i}$ (which
is the sum of the probabilities of each history that leads us to $I_{i}$...
under behavioral strategies $\beta $)

\pagebreak

\begin{equation}
\FRAME{itbphFU}{5.0004in}{4in}{0in}{\Qcb{Fig. 317.1}}{}{Figure}{\special%
{language "Scientific Word";type "GRAPHIC";maintain-aspect-ratio
TRUE;display "USEDEF";valid_file "T";width 5.0004in;height 4in;depth
0in;original-width 5.6886in;original-height 3.5201in;cropleft "0";croptop
"1";cropright "1";cropbottom "0";tempfilename
'JSH1WT0A.wmf';tempfile-properties "XPR";}}
\end{equation}

\subsubsection{Example:\ Consistency of beliefs at information sets on the
equilibrium path:}

In fig. 317.1 above, we decided that \{Unready,Acquiesce\} and \{Out,Fight\}
are both SPNE, although only \{Unready,Acquiesce\} is optimal at all
information sets (`continuation games').

Let's consider the (candidate) equilibrium \textbf{\{Unready,Acquiesce\}},
which is a \textit{SPNE}, is \textit{optimal at each information set,}

and potentially a \textit{WSE}.

The incumbent's information set \{Ready, Unready\} is thus \textbf{on} the
equilibrium path. \ 

Given the challenger's equilibrium strategy (if this is an equilibrium), the
only belief that is \textit{consistent} is $\Pr (Ready)=0,\Pr (Unready)=1.$

\pagebreak

\subsubsection{Example II:\ }

\[
\FRAME{itbphFU}{5.0004in}{4in}{0in}{\Qcb{Fig 326.1}}{}{Figure}{\special%
{language "Scientific Word";type "GRAPHIC";maintain-aspect-ratio
TRUE;display "USEDEF";valid_file "T";width 5.0004in;height 4in;depth
0in;original-width 5.1515in;original-height 4.8161in;cropleft "0";croptop
"1";cropright "1";cropbottom "0";tempfilename
'JSH1WT0B.wmf';tempfile-properties "XPR";}} 
\]

\textbf{1. }Let's examine the pairing of beliefs and strategies depicted
above. \ 

Player 1's strategy (E,J)\ is depicted in blue.

Player 2's strategy (G)\ is depicted in red. \ 

We already noted that each player's strategy is sequentially rational
(requirement 1).

Do player 2's beliefs meet the consistency requirement (requirement II)?

Yes, they do trivially, because player 2's information set is off the
equilibrium path, so any beliefs are permitted.

So the assesment that pairs the profile of strategies $\{E,J;G\}$ with
player 2's beliefs \ -- $\Pr (C)=\frac{2}{3}$,$\Pr (D)=\frac{1}{3}$ is a WSE.

\bigskip \pagebreak

\textbf{2. }What about the profile $\{D,J;G\}$ (see below)-- This is a NE,
and sequentially rational for the previously discussed belief system -- $\Pr
(C)=\frac{2}{3}$,$\Pr (D)=\frac{1}{3}$ .

But given 1's strategy this belief system is not consistent! \ 

Can this be a WSE for some consistent belief system?

\[
\FRAME{itbphFU}{5.0004in}{4in}{0in}{\Qcb{fig 326.1}}{}{Figure}{\special%
{language "Scientific Word";type "GRAPHIC";maintain-aspect-ratio
TRUE;display "USEDEF";valid_file "T";width 5.0004in;height 4in;depth
0in;original-width 5.3358in;original-height 4.6484in;cropleft "0";croptop
"1";cropright "1";cropbottom "0";tempfilename
'JSH1WT0C.wmf';tempfile-properties "XPR";}} 
\]

Given 1's strategy (D,J), the only consistent belief at information set
\{C,D\} is $\Pr (D)=1$.

But if $\Pr (D)=1$, 
\begin{eqnarray*}
EU_{2}(G) &=&0 \\
EU_{2}(F) &=&1
\end{eqnarray*}

So playing G is not sequentially rational!

\bigskip

\pagebreak

\textbf{Remember}:\ WSE\ is a \textit{refinement }of SPNE, which was a
refinement of NE. \ So, any WSE must be a SPNE and a NE. \ 

Let's examine a \textit{subset} of the strategic form, leaving out
strategies that include k (which is not subgame perfect).

\begin{tabular}{ccc}
& \textbf{F} & \textbf{G} \\ 
\textbf{C,J} & \underline{3},0 & 0,\underline{1} \\ 
\textbf{D,J} & 1,\underline{1} & \underline{2},0 \\ 
\textbf{E,J} & 2,\underline{0} & \underline{2},\underline{0}%
\end{tabular}

Since $\{E,J;G\}$ was the only SPNE in pure strategies we already knew it
was the only WSE in pure strategies.

\subsection{Signaling games}

Is your education valuable? Will it increase your income? \ Why?

\bigskip

Is it possible that education increases your income even though \textit{you
learn nothing of importance (other than this lecture), }

\textit{and emerge as ignorant and incompetent as you began? \ }

Perhaps education is only a tool employers use to separate the `good' from
`bad' workers.

\bigskip

If a `good' student can finish a degree at low cost, while a `bad' student
can do so only at a high cost ...

... \textit{good} students may obtain schooling and get good jobs (and find
the benefit worth the cost), because an employer knows that only a good
student would make this choice.

... while \textit{bad }students may decide not to go to school, even if
doing so would get them a better job.

This would be called a `\textbf{separating equilibrium}' in a \textbf{%
signaling game.}

\pagebreak

\subsubsection{Simple 2x2 signaling game (education and labor market)}

\textbf{Players: }Sender, Reciever, Nature

\bigskip

\textbf{Actions (and order):}

1. Nature: Good ability, Bad ability (picks with probabilities P$_{G}$ \& 1-P%
$_{G}$)

2. Sender: Signal Good (e.g, School), Signal Bad (e.g., Drink)

3. Reciever: Hire, Not hire

\bigskip

\textbf{Information structure:}

Sender observes ability

Reciever does not observe ability (Nature's move).

Receiver observes Sender's signal.

\bigskip

\textbf{Payoffs}

In general

U$_{Sender}$=Salary - cost of signal (function of ability)

U$_{Reciever}$= $\left\{ 
\begin{array}{c}
Production(function\text{ }of\text{ }ability)-Salary=\Pi \ \text{\ if hire}
\\ 
0\text{ if not hire}%
\end{array}%
\right. $

-- In general, the production function and wages are such that the reciever
will want to hire a good worker and not hire a bad worker.

-- However, if the receiver doesn't know a worker's ability, it would be
more profitable to never hire than to always hire.

\bigskip

\pagebreak

\begin{equation}
\FRAME{itbphF}{5in}{4in}{0in}{}{}{Figure}{\special{language "Scientific
Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display "USEDEF";valid_file
"T";width 5in;height 4in;depth 0in;original-width 7.3172in;original-height
4.9078in;cropleft "0";croptop "1";cropright "1";cropbottom "0";tempfilename
'JSH1WT0D.wmf';tempfile-properties "XPR";}}
\end{equation}

\subsubsection{Numerical Example:}

In a continuous-action example (see text) this will the worker will choose
the level of effort, the firm the level of pay for the good and bad jobs, and

\bigskip

Suppose the salary is set (exogenously)\ at $S=4$

The payoff to of not being hired is $2$ (second best job).

The production of a bad worker is $3$, so the firm `nets' -1 from hiring
such a worker.

The productivity of a good worker is $6$, so the firm `nets' 2 from hiring
such a worker.

The cost of sending a good signal (e.g., going to school)\ is 1 for the good
worker and 3 for the bad worker.

\bigskip

Finally, the probability of nature selecting a good worker is 0.1. \ 

This implies the firm would rather hire a good worker but not hire a bad
worker.

However, the firm would rather hire no one than hiring everyone.

\bigskip \pagebreak

If education means the difference between a job or no job, in this case,
only the good worker would choose to go to school (send a `good' signal).

However if education will not make a difference (say, you are hired in
either case), then neither type would go to school, as it is costly.

This yields the following game tree:

\begin{equation}
\FRAME{itbphF}{5in}{4in}{0in}{}{}{Figure}{\special{language "Scientific
Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display "USEDEF";valid_file
"T";width 5in;height 4in;depth 0in;original-width 7.8006in;original-height
5.7164in;cropleft "0";croptop "1";cropright "1";cropbottom "0";tempfilename
'JSH1WT0E.wmf';tempfile-properties "XPR";}}
\end{equation}

\textit{For example .... (check each outcome)}

If a worker is Good, goes to School and gets Hired, he will get a payoff of
3 (4 minus 1) and the firm gets a payoff of 2 \ This is the outcome in the
upper right corner.

\bigskip

To the left of this, if a worker is Good, goes to School and does Not get
hired, he will get a payoff of 1 (2 minus 1) and the firm gets a payoff of 2
\ 

\bigskip

To take another example, in the lower left outcome, if a worker is Bad,
chooses to Drink, and does Not get hired, he gets 2, and the firm gets zero.
\ 

Looking one to the left of this, if a worker is Bad, chooses to Drink, but
still gets Hired, he gets 4, and the firm gets -1. \ 

\bigskip

Let us check a potential `separating' equilibrium.

Suppose the sender plays (Drink,School) -- i.e., Drink if Bad, get an
education if Good.

Then suppose the reciever plays (Not, Hire) -- i.e., don't hire if Drink,
hire if School.

\bigskip

The only beliefs by a sender in such a situation that are \textit{consistent}
with this strategy is to believe Nature picked `Good' if he receives a Good
signal (School),

and Nature picked `Bad' if he receives a Bad signal (Drink). \ 

\bigskip

Given these beliefs and strategies, are these best responses at each
information set?

\bigskip

Compare EU's:

\[
EU_{S}(Drink,School;(Not,Hire))=0.1(4)+0.9(2)=0.2 
\]

Note this EU is \textit{additive} in the EU of each `type' (good or bad). \
\ Thus for this to be a BR it has to be a BR for each type of sender. \ Is
it?

\bigskip

Note that given the receiver's strategy (hire if and only if educated)\ the
decision to get an education is \textit{pivotal}. \ And as we set it up,
only the good type wants to get an education if it is pivotal. \ So each
type of sender is behaving optimally (i.e., the sender is behaving optimally
in each contingency)

\bigskip

Given the sender's strategy (get education if and only if Good), is the
reciever behaving optimally given his beliefs? \ Of course she is. \ 

As we set it up, the receiver prefers to hire the Good, and not hire the
bad. \ 

So this $\{Drink,School;Not,Hire\}$ is a WSE (sequential rational under the
beliefs mentioned above, which are of course consistent).

\bigskip

But this may not be the only WSE for this numerical example -- there may be
others. \ 

For example, pooling on no school if the employer will not hire for either
signal is a WSE.

I.e., \{$Drink,Drink;Not,Not$\} paired with the belief ($%
Pr(Good|Drink)=Pr(Good|School)=0.1$) is a WSE.

\textbf{Exercise: }Check this.

\pagebreak \bigskip

\subsubsection{Numerical Example II}

Let's change the payoffs.

Suppose the salary is set (exogenously)\ at $S=5$

The payoff to of not being hired is $0.$

The production of a bad worker is 4, so the firm `nets' -1 from hiring such
a worker.

The productivity of a good worker is 7, so the firm `nets' 2 from hiring
such a worker.

The cost of sending a good signal (e.g., going to school)\ is 1 for the good
worker and 2 for the bad worker.

\bigskip

Finally, the probability of nature selecting a good worker is 0.1. \ 

This implies (as before) the firm would rather hire a good worker but not
hire a bad worker.

However, (as before)\ the firm would rather hire no one than hiring everyone.

\bigskip

\textit{Notice:\ (Unlike before ) }If education means the difference between
a job or no job (i.e., is `pivotal'), both types would choose to go to
school (send a `good' signal).

However (as before) if education will not make a difference (say, you are
hired in either case), then neither type would go to school, as it is costly.

\pagebreak

This yields the following game tree:

\begin{equation}
\FRAME{itbphF}{5in}{4in}{0in}{}{}{Figure}{\special{language "Scientific
Word";type "GRAPHIC";maintain-aspect-ratio TRUE;display "USEDEF";valid_file
"T";width 5in;height 4in;depth 0in;original-width 7.2834in;original-height
5.7086in;cropleft "0";croptop "1";cropright "1";cropbottom "0";tempfilename
'JSH1WT0F.wmf';tempfile-properties "XPR";}}
\end{equation}

For example ....

\textbf{Exercise}: check each outcome

\bigskip

Let us check a potential `separating' equilibrium.

Suppose the sender plays (Drink,School) -- i.e., don't get an education if
bad, get an education if good.

Then suppose the reciever plays (Not, Hire) -- i.e., don't hire if Drink,
hire if education.

\bigskip

The only beliefs by a sender in such a situation that are \textit{consistent}
with this strategy is to believe Nature picked `Good' if he receives a Good
signal (education),

and Nature picked `Bad' if he receives a Bad signal (Drink). \ 

\bigskip

Given these beliefs and strategies, are these best responses at each
information set?\bigskip

Note that given the receiver's strategy (hire if and only if educated)\ the
decision to get an education is \textit{pivotal}. \ And as we set it up,
each type of sender wants to get an education if it is pivotal. \ So the
sender is not behaving optimally in the contingency where he is Bad (i.e.,
for this type). \ The bad type should \textit{also} want to pretend to be a
good type and get an education. .So this is not a WSE!

\bigskip

We can look for \textit{pooling} WSE in this case -- there seem to be no
separating WSE\ (in pure strategies) here.

Note that the firm prefers to hire no one than hire everyone (given the
setup). \ So, if both types do the same thing, the firm cannot tell which is
which, and will thus hire no one. \ So for any pooling equilibrium here, the
firm will hire no one. \ For example, we can look at \{$Drink,Drink;Not,Not$%
\}.

However, here the firm could assign \textit{any} beliefs if it observes a
good signal (education) -- such a signal is \textit{off the equilibrium
path. \ }It gets complicated now. \ One possible belief that the firm might
hold is that any player who signals \textit{Good} (gets an education) is a
Good worker. \ But if this were the beliefs, the firm would not be behaving
optimally -- it should thus \textit{hire} such a worker!

\bigskip

Another possible belief is that if a worker signals `Good' it has only a $.1$
probability of being a Good worker. \ I.e., the conditional probability
(after observing such off-path play) is the same as the unconditional
probability.This belief will make the strategy profile \{$%
Drink,Drink;Not,Not $\} sequentially rational. \ Thus \{$Drink,Drink;Not,Not$%
\} paired with the belief ($Pr(Good|Drink)=Pr(Good|School)=0.1$) is a WSE.

\bigskip

We can also look for other possibilities. \ We may find a separating (or
pooling) NE\ where the receiver plays \textit{behavioral strategies}, hiring
only some proportion of the time when she gets a good signal. \ In such a
case, the EU benefit to the sender may be such that getting an education is
only pivotal for the Good \ \ This is too advanced and complicated to cover
now.

\bigskip

\textit{This topic was tough, but it may still appear on the final! \ I will
not ask questions much above what was covered in class.}

\textit{Consider this part an 'introduction' to such games. \ }

\textit{\bigskip }

\textit{Next week we move on to finitely and infinitely repeated games. \
This will be a much easier topic, I think. SPNE will appear again and be
quite useful.}

\end{document}
