\section{Output inefficiency}\label{sec:ch10sec4}

A characteristic of perfect competition is that it secures an efficient allocation of resources when there are no externalities in the market: resources are used up to the point where their marginal cost equals their marginal value -- as measured by the price that consumers are willing to pay. But a monopoly structure does not yield this output. Consider Figure~\ref{fig:monopolyinefficiency}. 

% Figure 10.9 (called 10.7 in original text)
\input{figures/ch10/ch10fig9}

The monopolist's profit-maximizing output $Q_M$ is where $MC$ equals $MR$. This output is inefficient for the reason that we developed in Chapter~\ref{chap:welfare}: if output is increased beyond $Q_M$ the additional benefit exceeds the additional cost of producing it. The additional benefit is measured by the willingness of buyers to pay -- the market demand curve. The additional cost in that figure is the horizontal $MC$ curve under the assumption of constant returns to scale. Using the terminology from Chapter~\ref{chap:welfare}, there is a deadweight loss equal to the area ABF. This is termed \terminology{allocative inefficiency}.

\begin{DefBox}
\textbf{Allocative inefficiency} arises when resources are not appropriately allocated and result in deadweight losses .
\end{DefBox}

\subsection*{Perfect competition versus monopoly}

The area ABF can also be considered as the efficiency loss associated with having a monopoly rather than a perfectly competitive market structure. In perfect competition the supply curve is horizontal. This is achieved by having firms enter and exit when more or less must be produced. Accordingly, \textit{if the perfectly competitive industry's supply curve approximates the monopolist's long run marginal cost curve}\footnote{We could think of such a transformation coming from a single supplier taking over a number of small suppliers, and the monopolist would thus be a multi-plant firm. This nonetheless is quite an extreme possibility. A frequently cited reason for takeovers in the corporate world is that economies may materialize when firms are merged.}, we can say that if the monopoly were turned into a competitive industry, output would increase from $Q_M$ to $Q_{PC}$. The deadweight loss is one measure of the superiority of the perfectly competitive structure over the monopoly structure. With a linear demand and a horizontal $MC$ curve, the monopoly output is exactly one half of the competitive output. We will return to this in the next chapter.

Note that this critique of monopoly is not initially focused upon profit. While monopoly profits are what frequently irk the public, we have focused upon resource allocation inefficiencies. But in a real sense the two are related: monopoly inefficiencies arise through output being restricted, and it is this output reduction -- achieved by maintaining a higher than competitive price -- that gives rise to those profits. Nonetheless, there is more than just a shift in purchasing power from the buyer to the seller. Deadweight losses arise because output is at a level lower than the point where the $MC$ equals the value placed on the good; thus the economy is sacrificing the possibility of creating additional surplus.

Given that monopoly has this undesirable inefficiency, what measures should be taken, if any, to counter the inefficiency? We will see what Canada's Competition Act has to say in Chapter~\ref{chap:government} and also examine what other measures are available to control monopolies.