\section{Technology and innovation}\label{sec:ch10sec8}

Invention and innovation are critical aspects of the modern economy. Firms that cannot invent or innovate are liable to die. Invention is a genuine discovery, whereas innovation is the introduction of a new product or process. 

\begin{DefBox}
\textbf{Invention} is the discovery of a new product or process through research.

\textbf{Product innovation} refers to new or better products or services.

\textbf{Process innovation} refers to new or better production or supply.
\end{DefBox}

To this point we have said little that is good about monopolies: they cause a DWL and encourage wasteful rent-seeking behaviour. However, the economist Joseph Schumpeter (1883-1950) argued that, while monopoly leads to resource misallocation in the economy, this cost might be offset by the greater tendency for monopoly firms to invent and innovate. This is because such firms have more profit and therefore more resources with which to fund R\&D and may therefore be more innovative than competitive firms. If this were true then, taking a long-run dynamic view of the marketplace, monopolies could have lower costs and more advanced products than competitive firms and thus benefit the consumer.

While this argument has some logical appeal, it falls short on several counts. First, even if large firms carry out more research than competitive firms, there is no guarantee that the ensuing benefits carry over to the consumer. Second, the results of such research may be used to prevent entry into the industry in question. Firms may register their inventions and gain use protection before a competitor can come up with the same or a similar invention. Apple and Samsung each own tens of thousands of patents. Third, the empirical evidence on the location of most R\&D is inconclusive: a sector with several large firms, rather than one with a single or very many firms, may be best.  For example, if Apple did not have Samsung as a competitor, or vice versa, would the pace of innovation be as strong? These companies invent and innovate at a fantastic rate.

Fourth, much research has a `public good' aspect to it. Research carried out at universities and government-funded laboratories is sometimes referred to as basic research. This research explores the principles underlying chemistry, social relations, engineering forces, microbiology, etc. and has multiple applications in the commercial world. If disseminated, this research is like a public good -- its fruits can be used in many different applications, and its use in one area does not preclude its use in others. Consequently, rather than protecting monopolies on the promise of more R\&D, a superior government policy might be to invest directly in research and make the fruits of the research publicly available.

Modern economies have \terminology{patent laws}, which grant inventors a legal monopoly on use for a fixed period of time -- perhaps fifteen years. By preventing imitation, patent laws raise the incentive to conduct R\&D but do not establish a monopoly in the long run. Over the life of a patent the inventor charges a higher price than would exist if his invention were not protected; this both yields greater profits and provides the research incentive. When the patent expires, competition from other producers leads to higher output and lower prices for the product. Generic drugs are a good example of this phenomenon. 

\begin{DefBox}
\textbf{Patent laws} grant inventors a legal monopoly on use for a fixed period of time.
\end{DefBox}

The power of globalization once again is very relevant in patents. Not all countries have patent laws that are as strong as those in North America and Europe. The BRIC economies (Brazil, Russia, India and China) form an emerging power block. But their legal systems and enforcement systems are less well-developed than in Europe or North America. The absence of a strong and transparent legal structure inhibits research and development, because their fruits may be appropriated by competitors. 

At the same time, not every idea or new product can be registered for a patent. Evidently an idea for a TV show does not count. For example, Mark Burnett got the idea for the Survivor TV series from a similar product in Europe in the year 2000. This marked the start of reality TV in North America.  From this followed his Apprentice series, and imitators produced others: The Amazing Race, American Idol, So You Think You Can Dance, Dancing with the Stars, The Bachelor, The Biggest Loser, the X Factor, and so on. Furthermore, consider what might have happened if all North American TV watchers were subject to a monopoly content provider: we might have been denied all but a couple of such wonderful programs. 

\subsection*{Conclusion}

We have now examined two extreme types of market structure -- perfect competition and monopoly.  While many sectors of the economy operate in a way that is close to the competitive paradigm, very few are pure monopolies in that they have no close substitute products.  Even firms like Microsoft, or De Beers, that supply a huge percentage of the world market for their product would deny that they are monopolies and would argue that they are subject to strong competitive pressures from smaller or `fringe' producers. As a result we must look upon the monopoly paradigm as a useful way of analyzing markets, rather than being an exact description of the world. Accordingly, our next task is to examine how sectors with a few, several or multiple suppliers act when pursuing the objective of profit maximization. Many different market structures define the real economy, and we will concentrate on a limited number of the more important structures in the next chapter.