---
title: "Microeconomics (MSc)"
author: "Dr. David Reinstein, University of Exeter, [main web page](https://davidreinstein.wordpress.com), [innovationsinfundraising.org](https://innovationsinfundraising.org), Twitter: givingtools"
abstract: ""
cover-image: ""
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::gitbook:
    lib_dir: "book_assets"
    toc_depth: 2
    config:
      toc:
        before: |
          <li><a href="./">Microeconomics (MSc)<a></li>
        after: |
          <li><a href="https://bookdown.org" target="_blank">Published with bookdown</a></li>
        collapse: section
        scroll_highlight: yes
      fontsettings:
        theme: white
        size: 2
      sharing:
        facebook: yes
        twitter: yes
        google: no
        linkedin: yes
        weibo: yes
        instapaper: no
        vk: no
        #all: ['facebook', 'twitter', 'linkedin', 'weibo', 'instapaper']
    highlight: tango
    includes:
      in_header: [header_include.html]
    css: support/tufte_plus.css
    download: [pdf, epub, mobi]
    sharing:
      github: yes
      facebook: no
bookdown::pdf_book:
    keep_tex: yes
always_allow_html: yes
bibliography: ["support/reinstein_bibtex_from_dropbox.bib", "support/reinstein_references.bib", "support/packages.bib"] #this second one should be included in the citr settings, first comes from Zotero-dropbox sync, and it's re-downloaded below
biblio-style: apalike
link-citations: yes
github-repo: daaronr/micro-giving-pub
description: "Created for Exeter's BEEM101: Microeconomics, MSc module"
url: ''
tags: [Economics, Microeconomics, Applied, Theory, MSc, Textbook]
---


<!--
download: ["book.pdf", "book.epub"]
    includes:
      in_header:  style.css
-->

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

---
site: "bookdown::bookdown_site"
output:
  bookdown::gitbook:
    lib_dir: "book_assets"
  bookdown::pdf_book:
    keep_tex: yes
---

<!--
base file created from

`pandoc -f docx -t gfm -o writing_econ_gfm.md "bookoutline3-cutting examples down-cutnamesd.docx" `

and similar from

`pandoc -f docx -t gfm -o writing_econ_gfm1.md "Adapting back for BOOK --Ec831 outline-fillingindetails_forslides_edMiriam-conflict.docx"`

replacements needed:

- "\[ \]" surrounds math -- square brackets do not need 'escape' in main text
- colors need adjusting to 'format_with_col'

-->

Try downloading and accessing some functions here:
```{r download-formatting-to-support-folder}

library(here)
tryCatch(
  {
download.file(url = "https://raw.githubusercontent.com/daaronr/dr-rstuff/master/bookdown_template/support/header.html", destfile = here("support", "headerX.html"))
  download.file(url = "https://raw.githubusercontent.com/daaronr/dr-rstuff/master/bookdown_template/support/tufte_plus.css", destfile = here("support", "tufte_plusX.css"))
},  error = function(e) {
  print("you are not online, so we can't download")
}
)

tryCatch(
file.rename(here("support", "headerX.html"), here("support", "header.html"))
)

tryCatch(
file.rename(here("support", "tufte_plusX.css"), here("support", "tufte_plus.css"))
)

```

```{r download-fncns}

library(here)
library(devtools)

tryCatch(
  {
download.file(url = "https://raw.githubusercontent.com/daaronr/dr-rstuff/master/functions/functions.R", destfile = here("code", "functionsX.R"))
  download.file(url = "https://raw.githubusercontent.com/daaronr/dr-rstuff/master/functions/baseoptions.R", destfile = here("code", "baseoptionsX.R"))
},  error = function(e) {
  print("you are not online, so we source locally instead; hope you've updated")
  source(here("code", "functions.R")) # functions grabbed from web and created by us for analysis/output
 source(here("code", "baseoptions.R")) # Basic options used across files and shortcut functions, e.g., 'pp()' for print
}
)

#Note - workaround naming and code because otherwise a failed download seems to delete the destination file -- how to fix that? Renaming files?

file.rename(here::here("code", "functionsX.R"), here::here("code", "functions.R"))
file.rename(here::here("code", "baseoptionsX.R"), here::here("code", "baseoptions.R"))

source(here("code", "functions.R"))
source(here("code", "baseoptions.R"))

# Basic options used across files and shortcut functions, e.g., 'pp()' for print
# functions grabbed from web and created by us for analysis/output
```



```{r install,  echo = FALSE}
library(png)
library(grid)
```

```{r include=TRUE, eval=FALSE}
install.packages("webshot")
webshot::install_phantomjs()

```

```{r eval=FALSE}
install.packages("bookdown")


# or the development version
# devtools::install_github("rstudio/bookdown")
```

<!-- Download 'backup' bibtex file from dropbox (synced with Zotero), do the rename workaround -->


```{r }

tryCatch(
  {
download.file(url = "https://www.dropbox.com/s/3i8bjrgo8u08v5w/reinstein_bibtex.bib?raw=1", destfile = here("support", "reinstein_bibtex_from_dropboxX.bib"))


file.rename(here("support", "reinstein_bibtex_from_dropboxX.bib"), here("support", "reinstein_bibtex_from_dropbox.bib"))
  }
)

```

<!-- note installing packages is mainly removed because this is done in baseoptions.R -->

```{r }
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

```{r somefunctions}

#possibly move these to a separate file

#multi-output text color
#https://dr-harper.github.io/rmarkdown-cookbook/changing-font-colour.html#multi-output-text-colour
#We can then use the code as an inline R expression format_with_col("my text", "red")

format_with_col = function(x, color){
  if(knitr::is_latex_output())
    paste("\\textcolor{",color,"}{",x,"}",sep="")
  else if(knitr::is_html_output())
    paste("<font color='",color,"'>",x,"</font>",sep="")
  else
    x
}

```

```{r html, echo=FALSE}
# globally set chunk options
knitr::opts_chunk$set(fig.align='center', out.width='80%')

my_output <- knitr::opts_knit$get("rmarkdown.pandoc.to")

```


<!---
Can define text blocks here, refer to them again and again if desired
To add to ELE:
1. zip all content in '_book' folder
2. drag into VLE site
  - on prompt, create file resource
3. edit settings
  - right click and unzip
  - double-click starting file
  - set as main file
  - save
Now it is only accessible within the university login
-->


<!--chapter:end:index.Rmd-->

# Notes to students; overview {#note}

**Notes by instructor [David Reinstein](davidreinstein.wordpress.com/)**

## Focus and style of this module


**This is...**

1. a **Microeconomics** module.  We will learn and use maths and rigorous logic to consider deep questions of:

- Individual preferences, choices and optimization subject to constraints

- How individuals interact with one another, and the consequences of this (Game theory and mechanism design/agency problems)

- "Markets", prices, and "welfare" (how much value is created, who gets what, how do we consider these outcomes as a society)

- Psychological influences on the above, and limits to optimization (Behavioral Economics)

\


2. a  **Selective and bespoke** module

We will not cover 'all of Microeconomics'.  I will **focus on a limited set of concepts, models and application**s (and academic articles) that I hope you will find most interesting and useful.

We will skip many major important topics. But that's OK. We will learn tools, 'modes of thinking', and 'modes of proving results' that should strengthen your brain muscles and help you understand a much wider body of material.

- I often focus on models with 'surprising' or 'counter-intuitive' results. If you can explain the logic behind these (as well as the maths), you have achieved an understanding beyond a layman's intuition.

- I will try to incorporate a small set *applications* that are particularly relevant to your careers and your research interests.
,
<div class="marginnote">
I may also try to incorporate data, statistics/empirical modeling, and 'computer coding' because these will be super-important in your career.
</div>

*Please give me feedback on what you are interested in.*


3. an **Inquisitive module**: we ask **the 'how?', 'why?', 'so what?', and 'is this reasonable?'** ... of the models and concepts

When studying Economics, I often models presented, with strong assumptions, without any explanation of what these were for and how they could be justified. In this module we will always be skeptical and carefully ask 'what is the purpose' of each model and concept.

<div class="marginnote">
For example, 'what does it mean to say that something is a mixed-strategy Nash equilibrium?' ... Do we expect people to actually randomise in this way?
</div>



## Level of this module

This module is adapted to students from a variety of backgrounds.


For students interested in pursuing a PhD in Economics, and interested in research:

- We will cover some approaches and relevant building blocks, and I will point you to relevant materials for further enrichment

- I will incorporate content from academic literature ('articles in peer reviewed journals') and sophisticated research considerations

- I will incorporate my own research, research methods, and 'really getting our hands dirty' in digging into research questions, building theoretical (maths) models, and even perhaps experiments, coding, and data analysis.

- We will discuss *your* proposed research, relate microeconomic theory and models to this, and bring this into coursework

- I will try to help you 'think, research and write' like a good economist. (See my [Writing and research guide here](https://daaronr.github.io/writing_econ_research/about-this-work.html).

\

*However, please note that will not fully prepare you for doctoral study and academic research that requires advanced Microeconomics. MRes programmes and modules are better geared to this.*


\

For students coming from backgrounds with limited Economics training and limited abstract maths:

- I will explain most concepts 'from the ground up', or point you to what you will need to revise.

- I will present most material *also* at an intermediate level, using diagrams and simple intuition.

<div class="marginnote">
But note that you need to have SOME background in Economics to do well in this module. This is an MSc-level Economics module, after all.
</div>


\

For students who are not as fluent in English as they would like to be (perhaps many/most of our students):

- I will try to communicate clearly and at a reasonable pace, avoiding slang and using diagrams and maths to convey ideas

- Most of what I say in lectures/videos/tutorials is also in writing here in this web book (or in links)

- I will strongly encourage you to participate actively and try to help you improve your English communication skills.

- Interacting with  my [Writing and research guide here](https://daaronr.github.io/writing_econ_research/introduction.html) should also be helpful for this (and for your dissertation, if you will be writing one).





## Readings and resources {#readings}

This module will use a variety of resources: I will guide you through this via this online text. **Consider this online text the 'governing resource'.**

This draws heavily from several texts and other resources, **all of which are free and available online**. I will make it clear what come from where, and where you can find additional readings.

In particular:

1. "O-R": [Models in Microeconomic Theory, by Osborne and Rubinstein](https://www.openbookpublishers.com/product/1171)


> Our main aim is to give the reader an understanding of the concepts of model and equilibrium in microeconomic theory

This book is very precise, abstract (in a mathematical sense), and extremely well-written and full of wisdom and insight. It is *dense*; don 't worry if it takes you a long time to digest each concept (but if you get stuck, please do ask questions and try different approaches).  It is the most relevant (at least of these three) for further postgraduate/PhD study.

The authors claim 'the mathematics we use is elementary'. However, I think it contains a deep sort of maths that you will find challenging. As they say, 'many of the proofs involve sustained logical arguments'.


2. "QMC": [Quantum Microeconomics with Calculus](http://www.smallparty.org/yoram/quantum/quantumcalc.pdf) by Yoram Baumann, the self-proclaimed 'World's First Standup Economist'

This is somewhat less rigorous than O-R, but it has its own strengths. It approaches the material in an unusual way, and covers some material that is non traditionally covered in Economics courses (and skips many core concepts).

3. [Introduction to Economic Analysis version 2.1 by McAfee, Lewis, and Dale]](https://www.scholars.northwestern.edu/en/publications/introduction-to-economic-analysis-v21)

This is a more traditional text, with more 'computational-style' problems. It covers a wider set of Economics concepts than the others.


As I am incorporating some material *directly* from these works, I include the licences below (in the fold).

```{block2,  type='fold'}
Paste licenses here.

```

\

### Other relevant resources

**Simpler resources**

- Intermediate Microeconomics and its Application (an Undergraduate text) by Nicholson and Snyder. We have used this for a good chunk of this module in past years.

Free online access to this text may be available through your university/library (this works at Exeter) via [www.vlebooks.com](https://www.vlebooks.com/)

 - A free alternative, at a similar level: [Microeconomics an Open Text by Douglas Curtis and Ian Irvine](https://open.umn.edu/opentextbooks/textbooks/microeconomics-markets-methods-and-models)

-  Microeconomics by Hugh Gravelle and Ray Rees (version 3, widely available online for free):  A somewhat more advanced-intermediate-level text which I think you will find very useful; this is not yet officially Open-Source, but you can find it online for free (and the authors are OK with this, I have emailed them). Just do a web search for 'Gravelle and Rees Microeconomics' and you will find the pdf


\

**More rigorous, advanced and applied work will draw from**

- David Autor’s MIT Open Courseware “Microeconomic Theory and Public Policy”; [all the files for this course can be downloaded here)[https://ocw.mit.edu/courses/economics/14-03-microeconomic-theory-and-public-policy-fall-2016/lecture-notes/] this will give you a flavor of the PhD level micro
- David Autor. 14.03 Microeconomic Theory and Public Policy. Fall 2016. Massachusetts Institute of Technology: MIT OpenCourseWare, https://ocw.mit.edu. License: Creative Commons BY-NC-SA.


**Also**

For maths, see also [Mathematical methods for economic theory](https://mjo.osborne.economics.utoronto.ca/index.php/tutorial/index/1/toc) ... also free online!

\


- Academic articles and working papers as well as  some of my own writings... these are mentioned below




## This web-book: the Governing Resource

**Welcome to the exciting world of Microeconomics**

I wrote these notes to outline all of the material you are responsible for,  and to give you more resources and insight than I can convey in the short  amount contact time that we have.   I *don't* mean you're simply responsible for material in these printed notes:  I mean that these notes will aim to *refer to* everything that you are responsible for.

These notes are meant to be a complement to the texts and readings (and to the lectures/videos,  tutorials/problem sets and other material on the ELE) and to tie all of this together.

<!-- I am providing you these notes both as a complete file as well as divided by lecture week (approximately). Use either version.
-->

 I will try my best to stick to the structure outlined in these notes.  However, things may adjust slightly as we go and I will need to make updates to these notes as we go along.  I will try to keep you informed of any changes via the VLE forums.

<br> \bigskip

*Notes versus slides?:*

These notes should include all of the slide material *plus* explanatory notes and enrichment material. (However, I left out some of the jokes.) I use some abbreviations before certain notes to help you understand how to digest this.  I will also  make the slides themselves available even though they are redundant– these notes have everything.

<br> \bigskip


### 'Markers', abbreviations, and callouts/boxes {#markers}

- OR:  Osborne and Rubinstein's [Models in Microeconomic Theory]((https://www.openbookpublishers.com/product/1171)


- 'McDL' refers to the [McAfee, Dale and Lewis text](https://www.scholars.northwestern.edu/en/publications/introduction-to-economic-analysis-v2)

- 'QMC' refers to ['Quantum Microeconomics with Calculus'](http://www.smallparty.org/yoram/quantum/quantumcalc.pdf)  (Baumann)

- 'NS'  refers to the Nicholson and Snyder text (usually chapters/numbering from the 12th edition is referenced;  if you're using an earlier edition, make sure the content corresponds)

- GR refers to Gravelle and Rees' Microeconomics text
- 'DA' refers to [David Autor's MIT open courseware notes for Microeconomic Theory and Public Policy](https://ocw.mit.edu/courses/economics/14-03-microeconomic-theory-and-public-policy-fall-2016/lecture-notes/)



<!--
- 'LC' indicates a lecturer comment I may make in lecture that you don't need to worry about (I meant to cut these but I sometimes forget to)
-->

`r format_with_col("Text in blue usually refers to questions you should  try to answer on your own (and I sometimes provide suggested answers in the footnotes. Try to answer it yourself first.)","blue")`

`r format_with_col("Some text is given in gray to indicate that this is a side point or a discussion. (I'm moving most of this to footnotes.)","gray")`


- 'Comprehension Q:' A question you should be able to answer if you are following along.


\


Margin notes:

<div class="marginnote">
These reflect brief comments, hints, side points, or reference links.

</div>


\

Folding boxes:

I will occasionally put additional material in boxes that can be unfolded. Sometimes I do this with "question and answer" so you've an opportunity to consider the answer before peeking. I also will do this with some material that involves a side conversation, or material that is very tangential.

Look in the fold:

```{block2,  type='fold'}

Greetings from the fold! The weather is nice in here.
```

<!---
batch file: 'bash \$HOME/"Google Drive/exeter_teaching/be2024/compileslidesnotes.sh"'
-->

<br> \bigskip

### Tech details of this book ('do I call this 'collophon'?)

I created this in Rmd (the 'R languages' 'markdown'), including some latex and html code too, with features adapted from the 'Tufte' package.

This is all put together by  Knitr and Bookdown using the 'Pandoc' conversion tool into an Html file which I'm hosting via Github pages for now. For details on how this works, I have a guide (to how I work in a similarly constructed 'bookdown') [HERE](https://daaronr.github.io/ea_giving_barriers/bookdown-appendix.html).

I'll occasionally use/show some R code; you should consider getting proficient in R and developing other 'data-sciencey' tech skills (things like R, Python, Stata, git, vim, shell scripts...)

Let me know if you want to look 'under the hood'...

```{block2,  type='fold'}

All source material and code can be accessed [on my Github repo](https://github.com/daaronr/micro-giving-pub)\\\\\lp;[[


```



<!--chapter:end:preamble.Rmd-->

# Outline and description of module {#outline}

**Planned syllabus, coverage:**

*The planned module outline is below;  as noted, timings may need to be adjusted,  and so material may cut for time constraints.  Feel free to skip over the first time you read these notes; you may want to refer to this later. The relevant part of this outline will be repeated at the beginning of each 'section.'*

## Overview, broad discussion and background, Explaining the module, Economic models (and maths tools), 'empirical' evidence**  (1/2 week approximately) {-}

*This content will take roughly half a week to cover, but we will return to it later*


<!--  `r format_with_col("First problem set: Mathematical approaches, economic models, revision and warm-up; (to add: some practice MCQ's)","blue")`
I will put up slides and other resources to help you with this (and we will re-cover some of this when we apply it throughout the module).

-->

Main reading: this web book

\
**Choose some readings on 'what is Economics'**, 'what are the purpose of Economic models', etc.: The readings below are each optional, but you must read *at least some of these*


- [McDL'](https://www.scholars.northwestern.edu/en/publications/introduction-to-economic-analysis-v2)  Chapter 1. A very simple undergraduate discussion of 'what is Economics, what are models' etc. You probably already know this but take a look just in case.

- Friedman, Milton. "The methodology of positive economics." (1953): 259. *Classic but perhaps outdated*

- [Sugden, Robert. "Credible worlds: the status of theoretical models in economics." Journal of Economic Methodology 7.1 (2000): 1-31.](https://www.tandfonline.com/doi/pdf/10.1080/135017800362220)

- [Sugden, Robert; Sitzia, Stefania. "Implementing theoretical models in the laboratory, and what this can and cannot achieve."  Journal of Economic Methodology, Vol. 18, 2011, p. 323-343.](https://ueaeco.github.io/working-papers/papers/cbess/UEA-CBESS-11-08.pdf)


*We will return to the issues of 'what and why models' throughout the module*.

\


**Empirics, maths revision**

1.  Discussion of 'empirical identification': [The mixtape, Scott Cunningham](http://scunning.com/cunningham_mixtape.pdf) pp 18-22

2.  Math revision concepts in

- [DA lecture 3](https://ocw.mit.edu/courses/economics/14-03-microeconomic-theory-and-public-policy-fall-2016/lecture-notes/MIT14_03F16_lec3MathTools.pdf); we may not use all of this

- QMC 1.3 (some good stuff; don't worry too much about all the derivative rules)


<div class="marginnote">
Some of the maths you need will also be simultaneously covered in your Optimisation and Econometric modules, sometimes at a higher level.
</div>


### Interactive exercises  ('problem set'): {-}

- Discussion and brief writing/speaking exercise on 'what Economic models are for', examples of models. {-}

## Preferences and utility; Choice (1.5 weeks approximately) {-}

What can we reasonably say about 'what people prefer and choose'? How can we state this carefully and precisely? What are the implications of particular 'reasonable' assumptions? (How) can we consider this in terms of 'optimization'? How have Economists and decision-scientists considered these questions?

<div class="marginnote">
Digging deeply into the logic and justifications for these models of choice, and alternative frameworks, will be very relevant in considering Behavioral Economics (This is a specialty of Ariel Rubinstein, one of the O-R text's two authors, and also a research interest of mine). In fact this text works behavioral economics in from the beginning.

I also think it is relevant in building empirical models such as 'models of consumer choices and preferences' in ways that are relevant to (e.g.) Business-data-analytics consulting.
</div>

We approach this in an abstract and logically and mathematically-rigorous, but not highly 'computational' way. Even if you have taken Economics as an undergraduate, you may not have seen this approach.

This will take at least two weeks to cover meaningfully.

\


```{block2,  type='note'}

We will not always be so mathematically rigorous throughout the module. However, I want to at least give you some of these tools, and a 'flavour' of this approach, which characterises modern Microeconomics (and will be at the core of PhD modules).

```

<!--`-->

Main reading (largely integrated into web book):

- O-R Chapters 1-2 (we will focus on selected material as described in this web book, but feel free to read these chapters in their entirety if you find it interesting)

```{block2,  type='note'}

*Within O-R chapter 1, feel free to skip*  the proof of propositions 1.1 ('representing preferences by utility functions') and 1.2 ('Preference relation not represented by utility function')

```  


\

Alternative treatments for some of this material (read this if you need a different approach than the ones O-R are taking):

- NS Chapter 2 "Assumptions about preferences" (covers a limited subset in a very simple way)
\


Supplementary and conceptual readings (please read at least *some* of these):


- D. Colander \\Edgeworth's Hedonimeter and the Quest to Measure Utility'' Journal of Economic Perspectives, Spring 2007: 215-225.

- [Sugden, Robert. "Looking for a Psychology for the Inner Rational Agent." Social Theory and Practice 41, no. 4 (2015): 579-98.](http://www.jstor.org/stable/24575750).

- [‘Predicting Hunger: The Effects of Appetite and Delay on Choice’; Read and van Leeuwen, 1998](https://www.sciencedirect.com/science/article/pii/S0749597898928035) ... and related work... worth getting a sense of this, particularly in the context of the Sugden (2015) reading, such as [Kahneman and Richard H. Thaler. 2006. "Anomalies: Utility Maximization and Experienced Utility." Journal of Economic Perspectives, 20 (1): 221-234](https://pdfs.semanticscholar.org/2899/d2516fb877da40c4a7b95fe0b084a103b9bf.pdf)


Other/alternative readings:

- NS, chapter 2, p. 55-57 'assumptions about preferences' (a simpler brief and partial treatment)


### Interactive exercises ('problem set') {-}

1. Discussion and writing on the realism of particular assumptions and on 'observing preferences and observing violations of transitivity'

2. Preferences and utility: Definitions, proofs, and sketched examples -- Problems and exercises

Question 6 on the "Shepard scale" is very cool!
\


## Preferences under uncertainty and over time; 'Expected Utility';  risk-aversion, finance and diversification; time-preference and discounting (1 weeks approx.) {-}

With uncertainty (outcomes are unknown when choices are made, but we assume the *probabilities* of outcomes are known), again ...

> What can we reasonably say about 'what people prefer and choose'? How can we state this carefully and precisely? What are the implications of particular 'reasonable' assumptions? (How) can we consider this in terms of 'optimization'? How have Economists and decision-scientists considered these questions?

Applications (esp. to finance):

- How do we express and measure 'risk aversion'?

- How does this affect (financial, investment, insurance) choices?  Why does 'diversification improve outcomes for the risk-averse'?

- How does this affect asset prices (Stocks, bonds, etc.) with 'efficient markets' (this will be briefly defined)

- Behavioral Economics: the "Allais Paradox" (first treatment, time-permitting)


\

Brief (first) presentation: preferences (and optimization over) assets and consumption over time; 'discounting'

\

Main Readings:

*Web book*, and also...

- O-R Chapter 3: "Preferences under Uncertainty"

```{block2,  type='note'}

*Feel free to skip*  the proof of propositions 3.2 ('Continuity and independence implies EU'). It's interesting but we can't cover everything.

If you don't understand the proof of prop. 3.3. (the equivalence of risk aversion and the concavity of the Bernoulli function) that's also OK. 

```

- McDL: 13.3 on 'dynamic choice'



\

**Other resources/references**

- QMC: Ch 5, optimization and risk; selections (good discussion of diversification but doesn't use utility functions)

- DA: [Lecture Note 16: Uncertainty, Risk Preference, and Expected Utility Theory](https://ocw.mit.edu/courses/economics/14-03-microeconomic-theory-and-public-policy-fall-2016/lecture-notes/MIT14_03F16_lec16.pdf)

- NS: Ch 4 (not including 4a)

- McDL: Ch 13, section 4 'Risk Aversion'


\


Supplementary recommended readings:

- Holt, C., and S. Laury (2002), Risk Aversion and Incentive Effects, American Economic Review, v. 92 (5): 1644-1655.

- For a popular audience: Reinstein (2016) 'Should you hedge your bets on a Brexit?' [LINK](https://davidreinstein.wordpress.com/2016/06/19/should-you-hedge-your-bets-on-a-brexit/)


\

### Interactive exercises ('problem set')  {-}

- Proof-based problem

- Example-based problem (lotteries, insurance, investment)

- Allais Paradox question

- Potentially: data/code based simulation

- Discussion/explanation question

  \


## Consumer preferences, indifference curves/sets (0.5 weeks approx) {-}

- O-R Chapter 4, selected parts

<div class="marginnote">
Note this chapter covers only a simple 2-good example. If time permits, we will consider the properties of examples with "$N>2$ goods"; $N>2$ is necessary to consider reasonable substitution patterns between goods for many applications. You will also probably learn the N-good optimisation in your maths module.
</div>
 
```{block2,  type='note'}

*Within O-R chapter 4, feel free to skip*
  
- Section 4.6 "Differentiability" (just look at figure 4.5 to get the basic idea)

```

\

 - AUT: [Lecture 3 - Axioms of Consumer Preference and the Theory of
Choice](https://ocw.mit.edu/courses/economics/14-03-microeconomic-theory-and-public-policy-fall-2016/lecture-notes/MIT14_03F16_lec3.pdf) (note, slightly different notation)
    - Simpler treatment - NS: Ch 2


### Interactive exercises ('problem set')  {-}

- Standard problem set problems

- Business case study/investigation: modeling 'which (sorts of) new product will consumers prefer?'

\

## Consumer behavior/Individual (and market) demand functions and their properties (1 week approx.) {-}

- Formal definition
- Derivation of 'demand' from individual optimization
- 'Comparative statics': responses to changes in price, income, etc.;
    - Application: The mysterious 'Giffen good'
- Elasticities
    - Application: empirical models
- Aggregation and 'market demand'
- Welfare and consumer surplus (brief)


Main reading (largely integrated into web book):

- O-R Chapter 5 (selections)

```{block2,  type='note'}

Feel free to skip the first part of O-R section 5.5 on 'rationalizing a demand function.'

But make sure you *do read* the discussion of 'the weak axiom of revealed preferences' (WARP).

```  


\

Alternative treatments: 

- DA: [Lecture Note 6 – Demand Functions: Income Effects, Substitution Effects, and Labor Supply](https://ocw.mit.edu/courses/economics/14-03-microeconomic-theory-and-public-policy-fall-2016/lecture-notes/MIT14_03F16_lec6.pdf), esp. sections 1-3

- McDL:  3.1 (elasticity), parts of Ch 12; Warning: this book's treatment of consumer surplus seems to be incorrect (or at least vastly oversimplified) from my PoV

- QMC: Chapter 17 (elasticities -- very clear treatment, Chapter 18 (supply and demand details)

\


Supplementary recommended readings:

- Loomes, Graham, Chris Starmer, and Robert Sugden, 1991. "Observing Violations of Transitivity by Experimental Methods". [jstor link](https://www.jstor.org/stable/2938263?seq=1#metadata_info_tab_contents)

- Choi, Syngjoo, et al. "Who is (more) rational?." The American Economic Review 104.6 (2014): 1518-1550.a.

- Waldfogel, Joel. "The deadweight loss of Christmas." The American Economic Review 83.5 (1993): 1328-1336.

- [Reinstein, David. "The Economics of the Gift." (2014; working paper and chapter in Gift Giving and the'embedded' Economy in the Ancient World, F Carlà, M Gori - 2014).](https://core.ac.uk/download/pdf/74369736.pdf)

\

### Interactive exercises ('problem set')  {-}

- Something about finding departures from the WARP in real data?
\

- O-R problem 5 (introduces the 'expenditure function)
- O-R problem 8-9 (re-)introduce 'time preferences' 


### Noting a major "skip" {-}

**Note here that we are SKIPPING several key topics!**

- In the consumer behavior, we are skipping core topics including formal definitions of Hicksian demand, substitution and income effects, the 'and the Slutsky decomposition. 

<div class="marginnote">

These are important for characterizing the welfare effects of price changes and policy changes (including taxation), and for conceptualizing issues involving multiple constraints on consumer optimization (e.g., rationing). I will provide notes on these for anyone who is interested. 

</div>



\

We are also skipping most of the 'production side'

- Production functions, firm optimisation

- Perfect competition/competitive equilibrium; short and long run

- Supply and demand systems

- General equilibrium and it's welfare properties

- Market failures, including externalities/public goods (although these come back as examples in our Game Theory discussion)

*We this skip these chapters in the text(s). So skip O-R chapter 6 and skip any material on this in the other texts.*

<div class="marginnote">
 Or read it if you want to; you might find it useful in getting an overall picture, but it's up to you. 
</div>

To the extent that these are relevant to material we cover later, I will give you the 'short version' when it comes up.

\


## 'Monopolies and pricing of profit-maximizing price-setting firms' (especially monopolies); price discrimination/market segmentation (1.5 weeks approx.) {-}

How would a firm that knows all the stuff above about consumer demand set prices in order to maximise its profit? What if it can segment its customers into different ‘markets’ with distinct demand functions? What are the consequences of all of this for profit and welfare?

Today’s businesses, consultants, data analysts, and regulators are extremely interested in these questions. This is a very practical topic. We will consider it both from the theoretical optimization point view (while always gaining insights from the maths). We will also dig into “what do firms *actually* do in setting pricing” and “how can data be used to do this better?”

\

Main reading

- O-R ch. 7 "Monopoly"

We will focus on the (standard) profit-maximisation goal, although their discussion of other goals is interesting. 

They define consumer and producer surplus here to give a brief treatment of the deadweight loss of monopoly.

They also have a very interesting treatment of the two main types of price discrimination; 'Implicit discrimination' foreshadows our later study of asymmetric information and agency problems. 

<div class="marginnote">
There are various names of each of these ... I am familiar with "second degree  = self-selection" and "third degree or 'explicit segmentation'". They call the latter 'Implicit discrimination.
</div>
 

\

Alternative treatments:

- McDL: Ch 15 -- Monopoly
- NS: 11.2-11.4
- QMC: Appendix A.1 - Monopoly - the vanilla version

\
Supplementary readings:

- Article: [Should we help companies tailor prices to your wage packet?](https://theconversation.com/should-we-help-companies-tailor-prices-to-your-wage-packet-47719)
- With accompanying [worked examples](https://docs.google.com/document/d/16jos_PT9w1wGpyD5A8ZiWJ9HW6kaLOJH6EWc_AYWnkk/pub)
- More advanced: 'The Government May Want to Encourage Price Discrimination by Income' [Linked here](https://davidreinstein.wordpress.com/research-and-publications/)



### Interactive exercises ('problem set')  {-}

OR question 1, dbl marginalization (very relevant and interesting!)
OR question 5-7

Possible exercise involving data and 'choosing a firm's best price and segmentation'

<!--
- ** Imperfect competition (brief) ** (multiple firms with market power; esp Bertrand/Cournot models)
-->

## Noting a major "skip" {-}

Here we are largely skipping

- 'imperfect competition' and interactions between firms. We may come back to this briefly in our 'game theory' examples.



## Strategic interactions: Game theory (and evidence) (1.5 weeks approx) {-}

The outcomes we care about don't only depend on what *we choose*, but on what *others' choose* (and perhaps we can also influence the latter). How should we think about choices and outcomes in such settings? What will we predict will happen in single or repeated interactions, and what will affect these outcomes? 

\

- Strategic interactions
- Stating the game in 'normal' form
- Dominance, iterated dominance, Rationalisability
- Best response functions, Nash equilibrium in pure and mixed strategies 

<div class="marginnote">
(if time is tight we may skip mixed strategies)
</div>
 

- Extensive-form games; sequential aand repeated games 

- Backwards induction and Subgame Perfection (subgame-perfect Nash equilibrium)

- Infinitely or indefinitely repeated games

- Games of imperfect information

- Behavioral economics: transforming material payoffs into *true* payoffs
- Evidence on play in strategic situations, including in 'experiments'

- Time-permitting: Research applications to 'Marriage markets and Tinder', 'Fair trade and consumer altruism'

\


We will incorporate material from:

- O-R Chapters 15-16 
  - *Skip* sections 15.5 (Strictly competitive games), and 15.6 (Kantian equilibrium)
  - In chapter 16, we may or may not skip the formal depiction of extensive/sequential games; this depends on our timing

\
In addition to (some) of the Game Theory material covered in O-R: 

- We will also focus on 'Dominant strategies', and 'Iterated-strict-dominance'. Compared to 'Nash equilibria', these are (IMHO) more justifiable as *predictors of choices* especially in one-shot games. 

- We will consider games involving asymmetric information, especially 'signaling games'

\

Other readings to be incorporated:

- QMC: Ch 9, 10, 12 (includes 'iterated dominance')
- McDL: Ch 16
- NS: Ch 5 (parts)
- Several additional concepts from handout, plus supplement on experiments

\

Supplementary recommended reading:

- Spence, M. (1973): 'Job market signaling,'? The Quarterly Journal of Economics, vol. 87, pp. 355-374.

- 'Ten Little Treasures of Game Theory and Ten Intuitive Contradictions' Goeree and Holt, 2001

\


  

## Behavioural economics: Selected further concepts (1 weeks approx) {-}

- Overview: Limits to cognition, willpower, self-interest, applications

- Time-inconsistent preferences ('hyperbolic or $\beta$-$\delta$ discounting)

- Reference-dependent preferences and 'prospect theory' (also re-capping the Allais paradox)

- Evidence for these 'anomalies' 
  - Aside on  the nature of credible evidence, replication and open-science

- Modeling charitable giving and 'other-regarding behavior'; departures from and barriers to 'effective altruism'

\


Overall each reading is optional, but to do well on the exam you need to do at least some of these readings, and/or cover some of the more advanced material.

\

Supplementary reading: theory

(31 Jul 2020 note: This will be revised)

- Amos Tversky & Daniel Kahneman, 1979. "Prospect Theory: An Analysis of Decision under Risk" (*Seminal*)

\

Supplementary reading: applications and empirical work

- DellaVigna, Stefano. "Psychology and economics: Evidence from the field." Journal of Economic literature 47.2 (2009): 315-372.
- Benartzi, S. & Thaler, R.H., 2007. Heuristics and biases in retirement savings behavior. The journal of economic perspectives, pp.81-104.
- Farber, H., 2008. Reference-dependent preferences and labor supply: The case of New York City taxi drivers. The American Economic Review. Available at: http://www.ingentaconnect.com/content/aea/aer/2008/00000098/00000003/art00021 [Accessed November 19, 2015].

\
    
Other-regarding behavior and charitable giving: 

- Kellner, Reinstein and Riener, 2016. [Conditional generosity and uncertain income: Field and lab evidence](https://dl.dropboxusercontent.com/u/91553/Giving_and_Probability.pdf)

- Material linked at [giveifyouwin.org](giveifyouwin.org) and  [innovationsinfundraising.org](innovationsinfundraising.org)
- [Increasing effective charitable giving: The puzzle, what we know, what we need to know next](https://daaronr.github.io/ea_giving_barriers/outline.html)

## Recap, discussion of research, projects, and exams (1/2 week) {-}
    
## Supplement (optional): Asymmetric information (Moral hazard, adverse selection, signaling) and applications {-}

'Principal-agent' models to be integrated into lecture notes

- Moral hazard and incentives

- 'Lemons model'

- Hidden information about types/adverse-selection (time-permitting)

- 'Live research' application (time-permitting): Financial market micro-structure and government policy

\

Readings:

- O-R Chapter 14 on a market with Asymmetric Information (a highly formal treatment of the adverse-selection context)


<!--chapter:end:outline.Rmd-->

