\section{Land}\label{sec:ch12sec7}

Land is an input used in production, though is not a capital good in the way we defined capital goods earlier -- production inputs that are themselves produced in the economy. Land is relatively fixed in supply \textit{to the economy}, even in the long run. While this may not be literally true -- the Netherlands reclaimed from the sea a great quantity of low-lying farmland, and fertilizers can turn marginal land into fertile land -- it is a good approximation to reality. Figure~\ref{fig:marketlandservices} shows the derived demand $D_0$ for land services. With a fixed supply $S$, the equilibrium rental is $R_0$.

% Figure 12.9
\input{figures/ch12/ch12fig9} 

In contrast to this economy-wide perspective, consider now a retailer who rents space in a commercial mall. The area around the mall experiences a surge in development and more people are shopping and doing business there. The retailer finds that she sells more, but at the same time higher rents make it difficult for her to increase her profit. The reason why her profit may not increase greatly is that the demand for retail space is a \textit{derived demand}: with growth in the surrounding area many more retailers would like to rent space at the previous rent. Accordingly, the market demand curve for rental space shifts upwards, and this in turn leads to an increase in the price of rental space.

Of course, land has many uses and the returns to land must reflect this. Land in downtown Vancouver is priced higher than land in rural Saskatchewan. Land cannot be moved from the latter to the former location however, and therefore the rent differences represent an equilibrium. In contrast, land in downtown Winnipeg that is used for a parking lot may not be able to compete with the use of that land for office development. Therefore, for it to remain as a parking lot, the rental must reflect its high opportunity cost. This explains why parking fees in big US cities such as Boston or New York may run to \$40 per day.  If the parking owners could not obtain this fee, they could profitable sell the land to a developer. Ultimately it is the \textit{value in its most productive use} that determines the price of land.