\section{The politics of protection}\label{sec:ch15sec6}

Objections to imports are frequent and come from many different sectors of the economy. In the face of the gains from trade which we have illustrated in this chapter, why do we observe such strong rejections to imported goods and services?

\subsection*{Structural change and technology}

In a nutshell the answer is that while consumers in the aggregate gain from the reduction of trade barriers, and there is a net gain to the economy at large, some \textit{individual sectors of the economy lose out}. Not surprisingly the sectors that will be adversely affected are vociferous in lodging their objections. Sectors of the economy that cannot compete with overseas suppliers generally see a reduction in jobs. This has been the case in the manufacturing sector of the Canadian and US economies in the most recent two decades, as manufacturing and assembly has flown off-shore to Asia and Mexico where labour costs are lower. Domestic job losses are painful, and frequently workers who have spent decades in a particular job find reemployment difficult, and rarely get as high a wage as in their displaced job.

Such job losses are reflected in calls for tariffs on imports from China in order to `level the playing field' -- that is, to counter the impact of lower wages in China. Of course it is precisely because of lower labour costs in China that the Canadian consumer benefits. 

In Canada we deal with such dislocation first by providing `unemployment' payments to workers and second by furnishing retraining allowances from Canada's Employment Insurance program. Of course, such support does not guarantee an equally good alternative job. But structural changes in the economy, due to both internal and external developments, must be confronted. For example, the information technology revolution made tens of thousands of `data entry' workers redundant. Should producers have shunned the technological developments which increased their productivity dramatically? If they did, would they be able to compete in world markets?

While job losses feature heavily in protests against technological development and freer trade, most modern economies continue to grow and create more jobs in the service sector than are lost in the manufacturing sector. Developed economies now have many more workers in service than manufacture. Service jobs are not just composed of low-wage jobs in fast food establishments -- `Mcjobs', they are high paying jobs in the health, education, legal, financial and communications sectors of the economy.

\subsection*{Successful lobbying and concentration}

While efforts to protect manufacture have not resulted in significant barriers to imports of manufactures, objections in some specific sectors of the economy seem to be effective worldwide. One sector that stands out is agriculture. Not only does Canada have very steep barriers to imports of dairy products such as milk and cheese, the US and the European Union have policies that limit imports and subsidize their domestic producers. Europe has had its `wine lakes' and `butter mountains' due to excessive government intervention in these sectors; the US has subsidies to sugar producers and Canada has a system of quotas on the domestic supply of dairy products and corresponding limits on imports.

Evidently, in the case of agriculture, political conditions are conducive to the continuance of protection and what is called `supply management' -- domestic production quotas. The reason for `successful' supply limitation appears to rest in the geographic concentration of potential beneficiaries of such protection and the scattered beneficiaries of freer trade on the one hand, and the costs and benefits of political organization on the other: Farmers tend to be concentrated in a limited number of rural electoral ridings and hence they can collectively have a major impact on electoral outcomes. Second, the benefits that accrue to trade restriction are heavily concentrated in the economy -- keep in mind that about three percent of the population lives on farms, or relies on farming for its income. By contrast the costs on a per person scale are small, and are spread over the whole population. Thus, in terms of the costs of political organization, the incentives for consumers are small, but the incentives for producers are high.

In addition to the differing patterns of costs and benefits, rural communities tend to be more successful in pushing trade restrictions based on a `way of life' argument. By permitting imports that might displace local supply, lobbyists are frequently successful in convincing politicians that long-standing way-of-life traditions would be endangered, even if such `traditions' are accompanied by monopoly purchase -- as was the case under the Canadian Wheat Board until 2012, or tariffs as high as 250\% -- as is the case on cheese above a specific import quota. 

\subsection*{Valid trade barriers: infant industries and dumping?}

An argument that carries both intellectual and emotional appeal to voters is the `infant industry' argument. The argument goes as follows: new ventures and sectors of the economy may require time before that can compete internationally: scale economies may be involved for example, and time may be required for producers to expand their scale of operation, at which time costs will have fallen to international (i.e. competitive) levels. In addition learning by doing may be critical in more high-tech sectors, and once again, with the passage of time costs should decline for this reason also.

There are two problems with this stance: first, is that these infants would have insufficient incentive to grow up and become competitive; the second is why the private sector cannot invest sufficiently in promising young sectors.  A protection measure that is initially intended to be temporary can become permanent because of the potential job losses associated with a cessation of the protection to an industry that fails to become internationally competitive. Furthermore, employees and managers in protected sectors have insufficient incentive to make their production competitive if they realize that their government will always be there to protect them. As for the need for protection rather than private sector investment: if the industry is such a good idea in the long run, society should begin by asking why private firms cannot borrow the money to see them through the early period when they are losing out to more efficient foreign firms. If private lenders are not prepared to invest or lend, should the government be willing to step in? Does the government know more than the private sector about long-term investment returns?

In contrast to the infant industry argument, economists are more favourable to restrictions that are aimed to prevent `dumping'. \terminology{Dumping} is a predatory practice, based on artificial costs aimed at driving out domestic producers.

\begin{DefBox}
\textbf{Dumping} is a predatory practice, based on artificial costs aimed at driving out domestic producers.
\end{DefBox}

Dumping may occur either because foreign suppliers choose to sell at artificially low prices (prices below their marginal cost for example), or because of surpluses in foreign markets resulting from oversupply. For example, if, as a result of price support in its own market, a foreign government induced oversupply in butter and it chose to sell such butter on world markets at a price well below the going (`competitive') world supply price, such a sale would be a case of dumping. Alternatively, an established foreign supplier might choose to enter our domestic market by selling its products at artificially low prices, with a view to driving domestic competition out of the domestic market. Having driven out the domestic competition it would then be in a position to raise prices. Such behaviour differs from a permanently lower price on the part of foreign suppliers. This latter may be welcomed as a gain from trade, whereas the former may generate no gains and serve only to displace domestic labour and capital.