\section{Trading}\label{sec:ch3sec1}
The marketplace in today's economy has evolved from earlier times. It no longer has a unique form -- one where buyers and sellers physically come together for the purpose of exchange. Indeed, supermarkets require individuals to be physically present to make their purchases. But when purchasing an airline ticket, individuals simply go on line and interact with perhaps a number of different airlines (suppliers) simultaneously. Or again, individuals may simply give an instruction to their broker, who will execute a purchase on their behalf -- the broker performs the role of a middleman, who may additionally give advice to the purchaser. Or a marketing agency may decide to subcontract work to a translator or graphic artist who resides in Mumbai. In pure auctions (where a single work of art or a single residence is offered for sale) buyers compete one against the other for the single item supplied. These institutions are all different types of markets; they serve the purpose of facilitating exchange and trade. 

We should also keep in mind that not all goods and services in the modern economy are obtained through the marketplace. Schooling and health care are allocated in Canada primarily by government decree. In some instances the market plays a supporting role: Universities and colleges may levy fees, and most individuals must pay for their pharmaceuticals. In contrast, broadcasting services may carry a price of zero -- as with the Canadian Broadcasting Corporation.

The importance of the marketplace springs from its role as an allocating mechanism. Elevated prices effectively send a signal to suppliers that the buyers in the market place a high value on the product being traded; conversely when prices are low. Accordingly, suppliers may decide to cease supplying markets where prices do not remunerate them sufficiently, and redirect their energies and the productive resources under their control to other markets -- markets where the product being traded is more highly valued, and where the buyer is willing to pay more.

Whatever their form, the marketplace is central to the economy we live in. Not only does it facilitate trade, it also provides a means of earning a livelihood. Suppliers must hire resources -- human and non-human in order to bring their supplies to market and these resources must be paid a return -- income is generated. 

In this chapter we will examine the process of price formation -- how the prices that we observe in the marketplace come to be what they are. We will illustrate that the price for a good is inevitably linked to the quantity of a good; price and quantity are different sides of the same coin and cannot generally be analyzed separately. To understand this process more fully, we need to \textit{model} a typical market. The essentials are demand and supply.