\section{Market interventions}\label{sec:ch3sec7}

The freely functioning markets that we have developed certainly do not describe all markets. For example, minimum wages characterize the labour market, most agricultural markets have supply restrictions, apartments are subject to rent controls, and blood is not a freely traded market commodity in Canada. In short, price controls and quotas characterize many markets. \terminology{Price controls} are government rules or laws that inhibit the formation of market-determined prices. \terminology{Quotas} are physical restrictions on output. 

\begin{DefBox}
\textbf{Price controls} are government rules or laws that inhibit the formation of market-determined prices. 

\textbf{Quotas} are physical restrictions on output.
\end{DefBox}

Price controls come in the form of either \textit{floors} or \textit{ceilings}. 

\subsection*{Price ceilings}

Ceilings mean that suppliers cannot legally charge more than a specific price. Limits on apartment rents are one form of ceiling. In times of emergency -- such as flooding or famine, price controls are frequently imposed on foodstuffs, in conjunction with rationing, to ensure that access is not determined by who has the most income. The problem with price ceilings, however, is that they leave demand unsatisfied, and therefore they must be accompanied by some other allocation mechanism. 

Consider an environment where, for some reason -- perhaps a sudden and unanticipated growth in population -- rents increase. Let the resulting equilibrium be defined by the point $E_0$ in Figure~\ref{fig:priceceiling}. If the government were to decide that this is an unfair price because it places hardships on low- and middle-income households, it might impose a price limit, or ceiling, of $P_c$. The problem with such a limit is that excess demand results: Individuals want to rent more apartments than are available in the city. In a free market the price would adjust upward to eliminate the excess demand, but in this controlled environment it cannot. So some other way of allocating the available supply between demanders must evolve.

% Figure 3.6
\input{figures/ch3/ch3fig6}

In reality, most apartments are allocated to those households already occupying them. But what happens when such a resident household decides to purchase a home or move to another city? It holds a valuable asset, since the price/rent it is paying is less than the free-market price. Rather than give this surplus value to another family, it might decide to sublet at a price above what it currently pays. While this might be illegal, the family knows that there is excess demand and therefore such a solution is possible. A variation on this outcome is for an incoming tenant to pay money, sometimes directly to an existing tenant or to the building superintendent, or possibly to a real estate broker who will ``buy out'' existing tenants. This is called ``key money.'' 

Rent controls are widely studied in economics, and the consequences are well understood: Landlords tend not to repair or maintain their rental units and so the residential stock deteriorates. Builders realize that more money is to be made in building condominium units, or in converting rental units to condominiums. The frequent consequence is a \textit{reduction} in supply and a reduced quality. Market forces are hard to circumvent because, as we emphasized in Chapter~\ref{chap:intro}, economic players react to the incentives they face. This is an example of what we call the \textit{law of unintended consequences}.

\subsection*{Price floors}

An effective price floor sets the price \textit{above} the market-clearing price. A minimum wage is the most widespread example in the Canadian economy. Provinces each set their own minimum, and it is seen as a way of protecting the well-being of low-skill workers. Such a floor is illustrated in Figure~\ref{fig:pricefloor}. The free-market equilibrium is again $E_0$, but the effective market outcome is the combination of price and quantity corresponding to the point $E_f$ at the price floor, $P_f$. In this instance, there is excess supply equal to the amount $E_f$C.

% Figure 3.7
\input{figures/ch3/ch3fig7}

Note that there is a similarity between the outcomes defined in the floor and ceiling cases: The quantity actually traded is \textit{the lesser of the supply quantity and demand quantity at the going price: the short side dominates.}

\subsection*{Quotas}

A quota represents the right to supply a specified quantity of a good to the market. It is a means of keeping prices higher than the free-market equilibrium price. As an alternative to imposing a price floor, the government can generate a high price by restricting supply. 

Agricultural markets abound with examples. In these markets, farmers can supply only what they are permitted by the quota they hold, and there is usually a market for these quotas. For example, in several Canadian provinces it currently costs in the region of \$30,000 to purchase a quota granting the right to sell the milk of one cow. The cost of purchasing quotas can thus easily outstrip the cost of a farm and herd. Canadian cheese importers must pay for the right to import cheese from abroad. Restrictions also apply to poultry. The impact of all of these restrictions is to raise the domestic price above the free market price.

In Figure~\ref{fig:quota}, the free-market equilibrium is at $E_0$. In order to raise the price above $P_0$, the government restricts supply to $Q_q$ by granting quotas, which permit producers to supply a limited amount of the good in question. This supply is purchased at the price equal to $P_q$. 

% Figure 3.8
\input{figures/ch3/ch3fig8}

\subsection*{Modelling market interventions}

To illustrate the impact of these interventions on our numerical market model for natural gas, let us suppose that the government imposes a minimum price of \$6 -- above the equilibrium price obviously. We can easily determine the quantity supplied and demanded at such a price. On the supply side:

\begin{equation*}
P=1+(1/2)Q.
\end{equation*}

Hence at $P=6$ it follows that $6=1+(1/2)Q$; that is $5=(1/2)Q$. Thus $Q$ must take a value of 10, which is to say that suppliers \textit{would like to supply} 10 units at this price.

Correspondingly on the demand side:

\begin{equation*}
P=10-Q,
\end{equation*}

At $P=6$, it follows that $6=10-Q$; that is $Q=4$. So buyers \textit{would like to buy} 4 units at that price: there is excess supply. But we know that the short side of the market will win out, and so the actual amount traded at this restricted price will be 4 units.